---
title: "Women's Soccer Data"
output: html_notebook
---

We'll use this notebook to explore some NCAA women's soccer data and [our questions](https://docs.google.com/spreadsheets/d/1kTSrnl1URmcexIWLDMWqDnp3q-c2OS45uifnhjp10n4/edit#gid=0). First, let's load the tidyverse library:

```{r}
library(tidyverse)
```

We'll want to start with 2023 match data, which is located in [this GitHub repository](https://github.com/dwillis/NCAAWomensSoccerData). We can load the match data directly from GitHub using `read_csv`:

```{r}
read_csv("C:/Users/benst/Downloads/ncaa_womens_soccer_matchstats_2024.csv")
```

Let's take a look at this data to see what it contains, using `head` and `glimpse`:

```{r}
head(matches_2024)
glimpse(matches_2024)
```

Now, onto the questions using aggregates, which means `group_by` and `summarize`:

1.  How many shots in total has Maryland attempted and how many have been on goal? Calculate a percentage of shots on goal.

48.9 % of shots on goal. 22 on goal out of 45

```{r}
library(readr)



data <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2024.csv")


head(data)


matches_2024 |>
  group_by(team, sh_att, so_g) |>
  summarise(
    totalshots = sum(sh_att),
    shotongoal = sum (so_g),
    percentongoal = (shotongoal/totalshots)
  )
  



```

2.  What about the same shot stats from last season's first four games? You'll need to load the matches from 2023 - just change the season in the URL - and filter by date to find the first four games.

```{r}
library(readr)



data <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2023.csv")




```

3.  What about draws - Maryland started 2022 with 4 ties. Have any other teams had as many this season?

```{r}
```

4.  How many total corner kicks has Maryland had so far and how many have its opponents have? Calculate the difference between the two totals for every team.

```{r}
```