```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
Stats <- read_csv("https://raw.githubusercontent.com/dwillis/dwillis.github.io/refs/heads/main/docs/sports-data-files/footballlogs1124.csv")
```

I decided to study whether or not teams who run more than pass, win more games in college football. I have the necessary data. However, I will need to determine what outliar exist, including teams like Navy who run the option. I'm not sure how yet, but I need to determine an exact thing I'm looking for in these stats. For example, I need a way to determine how much more a team runs than throws to truly consider them in this field. According to TeamRankings.com, 82 teams run 50% or more of its plays. This is good because it means I will have a lot of data to work with. All I need to do is remove outliars such as Army, Navy and Airforce, who all run over 70% of the time.

```{r}
running_stats <- Stats |>
  group_by(TeamFull, Conference) |>  
  summarise(total_rushes = sum(RushingAtt, na.rm = TRUE),
            total_plays = sum(OffensivePlays, na.rm = TRUE),
            .groups = 'drop') |>  
  mutate(run_percentage = total_rushes / total_plays * 100) |>
  filter(run_percentage > 50) |>
  arrange(desc(run_percentage))  

print(running_stats)





```
```{r}
sec_stats <- running_stats |>
  filter(Conference == "Southeastern Conference")



```


```{r}
ggplot() +
  geom_bar(
    data = sec_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in SEC",
       x = "School",
       y = "Rushing Percentage")



```
Surprise, the Auburn Tigers are the only team in the SEC to run the ball over 60% of its plays.

```{r}
big_stats <- running_stats |>
  filter(Conference == "Big Ten Conference")

```
```{r}
ggplot() +
  geom_bar(
    data = big_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in B1G",
       x = "School",
       y = "Rushing Percentage")

```

For teams that run over 60% of the time, seeing Wisconsins and Minnesota on this list is kind of surprising. We should keep an eye on this.

```{r}
aac_stats <- running_stats |>
  filter(Conference == "American Athletic Conference")
```

```{r}
ggplot() +
  geom_bar(
    data = aac_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in AAC",
       x = "School",
       y = "Rushing Percentage")
```
This is a potential weakness here in the data. Navy and Army, both in the AAc, rush as much as 80% of its plays. This makes them outliars in this data. This makes my question a little tougher to answer, especially this year because both these teams are a combined 10-0.


```{r}
mwc_stats <- running_stats |>
  filter(Conference == "Mountain West Conference")
```

```{r}
ggplot() +
  geom_bar(
    data = mwc_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in Mountain West",
       x = "School",
       y = "Rushing Percentage")
```
Here's another example of a challenge presented with the Air Force Academy. Somehow, we need to find a way to distinguish these schools that run the option into their own category. However, New Mexico high percentage I find very interesting. All in all, the Mountain West is really interesting because every team is run heavier. I might explore this conference more.

Overall, I created bar graphs for conferences I found important in this data. I'm looking at how running translates to results, so that will be my next step in the process. However, power conferences such as the SEC and B1G are a lot less run heavy than mid major schools. This provides a challenge for this question because it begs the question; Is it because these mid majors simply lack aerial talent? Is this the standard for the conference? Another challenge is finding true correlation in this data. I will need to use the coordination and z score tools in R to really discover where the correlation resides in this data. I'm excited to explore more, but these charts do provide a good visual representation of the difference between some of these small conferences and the big powerhouses.
