```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
Stats <- read_csv("https://raw.githubusercontent.com/dwillis/dwillis.github.io/refs/heads/main/docs/sports-data-files/footballlogs1124.csv")
```

I decided to study whether or not teams who run more than pass, win more games in college football. I have the necessary data. However, I will need to determine what outliar exist, including teams like Navy who run the option. I'm not sure how yet, but I need to determine an exact thing I'm looking for in these stats. For example, I need a way to determine how much more a team runs than throws to truly consider them in this field. According to TeamRankings.com, 82 teams run 50% or more of its plays. This is good because it means I will have a lot of data to work with. All I need to do is remove outliars such as Army, Navy and Airforce, who all run over 70% of the time.

```{r}
running_stats <- Stats |>
  group_by(TeamFull, Conference) |>  
  summarise(total_rushes = sum(RushingAtt, na.rm = TRUE),
            total_plays = sum(OffensivePlays, na.rm = TRUE),
            .groups = 'drop') |>  
  mutate(run_percentage = total_rushes / total_plays * 100) |>
  filter(run_percentage > 50) |>
  arrange(desc(run_percentage))  

print(running_stats)





```
```{r}
sec_stats <- running_stats |>
  filter(Conference == "Southeastern Conference")



```


```{r}
ggplot() +
  geom_bar(
    data = sec_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in SEC",
       x = "School",
       y = "Rushing Percentage")



```
Surprise, the Auburn Tigers are the only team in the SEC to run the ball over 60% of its plays.

```{r}
big_stats <- running_stats |>
  filter(Conference == "Big Ten Conference")

```
```{r}
ggplot() +
  geom_bar(
    data = big_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in B1G",
       x = "School",
       y = "Rushing Percentage")

```

For teams that run over 60% of the time, seeing Wisconsins and Minnesota on this list is kind of surprising. We should keep an eye on this.

```{r}
aac_stats <- running_stats |>
  filter(Conference == "American Athletic Conference")
```

```{r}
ggplot() +
  geom_bar(
    data = aac_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in AAC",
       x = "School",
       y = "Rushing Percentage")
```
This is a potential weakness here in the data. Navy and Army, both in the AAc, rush as much as 80% of its plays. This makes them outliars in this data. This makes my question a little tougher to answer, especially this year because both these teams are a combined 10-0.


```{r}
mwc_stats <- running_stats |>
  filter(Conference == "Mountain West Conference")
```

```{r}
ggplot() +
  geom_bar(
    data = mwc_stats,
    aes(x = reorder(TeamFull, run_percentage),
        y = run_percentage),
    stat = "identity") +
  coord_flip() +
  labs(title = "Over 50% Rushing Teams in Mountain West",
       x = "School",
       y = "Rushing Percentage")
```
Here's another example of a challenge presented with the Air Force Academy. Somehow, we need to find a way to distinguish these schools that run the option into their own category. However, New Mexico high percentage I find very interesting. All in all, the Mountain West is really interesting because every team is run heavier. I might explore this conference more.

Overall, I created bar graphs for conferences I found important in this data. I'm looking at how running translates to results, so that will be my next step in the process. However, power conferences such as the SEC and B1G are a lot less run heavy than mid major schools. This provides a challenge for this question because it begs the question; Is it because these mid majors simply lack aerial talent? Is this the standard for the conference? Another challenge is finding true correlation in this data. I will need to use the coordination and z score tools in R to really discover where the correlation resides in this data. I'm excited to explore more, but these charts do provide a good visual representation of the difference between some of these small conferences and the big powerhouses.


```{r}
agg_stats <- running_stats |>
  group_by(Conference)|>
  summarize(avg_run_percentage = mean(run_percentage))

```

```{r}
ggplot() +
  geom_bar(
    data = agg_stats,
    aes(x = reorder(Conference, -avg_run_percentage), 
        weight = avg_run_percentage))  +
  labs(title = "Rushing Percentage Over Time by Conference (2011-2024)",
       x = "Conference",
       y = "Average Rushing Percentage")+
        
  theme_minimal() +
coord_flip()

```
Looking at this chart is very interesting because it highlights all the conferences in our data, and not just the 4 I looked at earlier. Every conference is over 50%, so what's interesting about this? Only 2 conferences are over 60%. It's Independent and the Mountain West. I saw that the Mountain West was interesting in my earlier studies, but now we see it on a scale with every other conference, including mid majors. I had wondered if smaller conferences simply ran more than the power conferences, but this chart suggest that maybe the Mountain West is simply different than most conferences. I'm glad this data shows the independents because that includes those schools like Navy, Army and Air force. However, Notre Dame is included, which drives the percentage down. That's important to look at here. 

```{r}
passing_stats <- Stats |>
  mutate(pass_percentage = (PassingAtt / (PassingAtt + RushingAtt)) * 100,
         rush_percentage = (RushingAtt / (PassingAtt + RushingAtt)) * 100)
```


```{r}
# Calculate yards per attempt, TD/INT ratio, and completion percentage
pass_heavy_teams <- passing_stats |>
  filter(pass_percentage > 50) |>
  mutate(
    yards_per_att = PassingYds / PassingAtt,  # Calculate yards per attempt
    td_int_ratio = ifelse(Interceptions == 0, NA, PassingTD / Interceptions),  # Handle division by zero
    completion_pct = (PassingCmp / PassingAtt) * 100  # Calculate completion percentage
  )

```

Now we have some data about pass heavy teams. There are quiet a lot of household names on here. Let's find a way to view how efficient teams are by conference.

```{r}
pass_eff <- pass_heavy_teams %>%
  group_by(Conference, Season) %>%
  summarize(
    avg_yards_per_att = mean(yards_per_att, na.rm = TRUE),
    avg_td_int_ratio = mean(td_int_ratio, na.rm = TRUE),
    avg_completion_pct = mean(completion_pct, na.rm = TRUE))
```
```{r}
pass_eff_long <- pass_eff |>
  pivot_longer(cols = c("avg_yards_per_att", "avg_td_int_ratio", "avg_completion_pct"),
               names_to = "Metric",
               values_to = "Value")
print(head(pass_eff_long))
```

Let's visualize this data to find how efficent each conference was with pass heavy teams.

```{r}
ggplot(pass_eff_long, aes(x = Season, y = Value, fill = Conference)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Offensive Efficiency for Pass-Heavy Teams (2011-2024)",
       x = "Season",
       y = "Efficiency Metric",
       fill = "Conference") +
  facet_wrap(~ Metric, scales = "free_y") +  
  theme_minimal() +
  theme(legend.position = "right")
```
As shown, I've tried to visualize the offensive metrics for pass heavy teams. However, the data is too much for the visualization to be easily understandable.Let's try another way.
```{r}
agg_eff <- pass_eff_long %>%
  group_by(Conference, Metric) %>%
  summarize(avg_value = mean(Value, na.rm = TRUE))
```
```{r}
ggplot(agg_eff, aes(x = reorder(Conference, -avg_value), y = avg_value, fill = Conference)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Offensive Efficiency for Pass-Heavy Teams (2011-2024)",
       x = "Conference",
       y = "Average Efficiency Metric",
       fill = "Conference") +
  facet_wrap(~ Metric, scales = "free_y") +  # Separate by each metric
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))
```
This is really interesting to look at it. According to the charts, The Pac-12 conference has been the best aerial conference since 2011, based on these metrics. The Independent conference is last, which is no surprise because those teams - Notre Dame in this case, don't throw very often. We noticed earlier how the Mountain West ran the ball over 60% of the time as a conference, so to see them on the lower side of these 3 graphs isn't surprising. The SEC and Pac-12, which have seen a lot of success in the passing game over the years, have also seen success period in the win column. You can make a strong argument here that for conferences like those, they rely more on passing because of the talent they possess at those skill positions. One thing I notice interesting here is the Sun Belt Conference. They are top 5 in all these categories. I wonder why. 

Also, how about the Big Ten?? The Big Ten is very low in efficiency in the passing game compared to other conferences.The Big ten runs the ball 56% of the time, which is somewhat higher, but not overwhelming. Are teams like Iowa and Nebraska bringing these numbers way down in the passing game?








```{r}
team_rushing_trends <- Stats |>
  mutate(Year = year(ymd(Date)),  
         RushingPct = RushingAtt / (PassingAtt + RushingAtt)) |>  
  group_by(TeamFull, Year) |>
  summarise(run_percentage = mean(RushingPct, na.rm = TRUE))
```
```{r}
team_rushing_pivot <- team_rushing_trends |>
  pivot_wider(names_from = Year, values_from = run_percentage)
```

```{r}
str(team_rushing_pivot)
```


```{r}
library(dplyr)

# Assuming 'team_rushing_pivot' has columns for each year (e.g., 2011, 2012, 2013, etc.)
# Step 1: Calculate the year-over-year percentage change for each team
team_rushing_pivot <- team_rushing_pivot %>%
  mutate(across(starts_with("20"), ~ (. - lag(.)) / lag(.) * 100, .names = "PctChange_{col}"))

# Step 2: Display the results
head(team_rushing_pivot)

```


```{r}
percentage_change <- team_rushing_trends |>
  mutate(across(starts_with("20"), ~ (. - lag(.)) / lag(.) * 100, .names = "PctChange_{col}"))

```



```{r}
library(dplyr)


library(dplyr)

# Assuming your data is named 'team_rushing_data'
# Step 1: Group by Team and arrange by Year to calculate year-over-year percentage change
team_rushing_changes <- team_rushing_trends %>%
  group_by(TeamFull) %>%
  arrange(Year) %>%  # Ensure the data is arranged by year within each team
  mutate(
    PctChange_RushingPct = (run_percentage - lag(run_percentage)) / lag(run_percentage) * 100
  ) %>%
  ungroup()

# Step 2: Replace any NA values in the percentage change with 0 (e.g., for the first year or where no lag exists)
team_rushing_changes <- team_rushing_changes %>%
  mutate(PctChange_RushingPct = replace_na(PctChange_RushingPct, 0))

# Step 3: Print the results to check the year-over-year percentage changes
print(head(team_rushing_changes))



```
```{r}
library(tidyr)

team_rushing_wide <- team_rushing_changes %>%
  select(-run_percentage) %>%  # Remove the 'run_percentage' column
  pivot_wider(
    names_from = Year,                  # Use 'Year' as the new column names
    values_from = PctChange_RushingPct   # The values will come from the year-over-year percentage change
  ) %>%
  select(-`2011`) %>%  # Remove the 2011 column as it will be NA
  mutate(TotalPctChange = rowSums(across(starts_with("20")), na.rm = TRUE))  # Calculate the total percentage change

# Step 2: Print the result with totals
print(head(team_rushing_wide))


```


```{r}
top_ten <- team_rushing_wide %>%
  arrange(desc(TotalPctChange)) %>%
  head(10)
```


```{r}
top_changers <- top_changers %>%
  filter(!is.na(TotalPctChange) & !is.na(TeamFull))

print(top_changers)

# Generate the bar plot
ggplot(top_ten, aes(x =reorder (TeamFull, TotalPctChange), weight=TotalPctChange)) +
  geom_bar()+
  labs(title = "Top 10 Teams with Greatest Rushing Percentage Change since 2011",
       x = "Team",
       y = "Total Percentage Change") +
  theme_minimal() +
  coord_flip()  # Flip coordinates for better readability
```
Circling back to what we know from earlier, the Pac-12 conference has been the heaviest throwing conference in recent history. With that being said, it's unbelievable the top 2 teams that have run the most in this span actually come from within the conference.

Also, if you look down just a bit, the former Pac-12 Arizona State Sun Devils also crack the top 6. Although the majority of teams in this conference have passed in the past decade. These 3 schools have widely deviated from their peers.

Let's put this into perspective with the rest of the conferences and how rushing trends have changed over time.
`
```{r}
# Calculate average rushing percentage by Conference and Year
rushing_trends <- Stats |>
  mutate(Year = year(ymd(Date)),
         RushingPct = RushingAtt / (PassingAtt + RushingAtt) * 100) |>
  group_by(Conference, Year) |>
  summarise(avg_rushing_pct = mean(RushingPct, na.rm = TRUE))

# Calculate overall average rushing percentage across all conferences by year
average_rushing <- rushing_trends |>
  group_by(Year) |>
  summarise(mean_rushing_pct = mean(avg_rushing_pct, na.rm = TRUE))

# Plot the data
ggplot() + 
  geom_line(data = rushing_trends, aes(x = Year, y = avg_rushing_pct, group = Conference), color = "grey") +  # Grey background for all conferences
  geom_line(data = filter(rushing_trends, Conference == "Pac-12 Conference"), aes(x = Year, y = avg_rushing_pct), color = "red") +  # Highlight Pac-12 in red
  
  geom_line(data = average_rushing, aes(x = Year, y = mean_rushing_pct), color = "black") +  # Add average rushing percentage in black
  labs(title = "Pac-12 reveals as one of the lowest rushing conferences",
       x = "Year",
       y = "Average Rushing Percentage") +
  scale_y_continuous(limits = c(0, 80)) +  # Adjust y-axis limits if needed
  theme_minimal()

```

When compared to all other conferences, the Pac-12 runs less than most. Despite this fact, and the massive drop off between 2017 and 2020, the top-two teams with the greatest percentage change in rushing percentage in the last 10 years are Washington State and Oregon State.

As indicated with the median average black line, The Pac-12 has never even been average overall in terms of rushing percentage. It's unreal that 3 teams from the conference saw the greatest increase in rushing since 2011.


let's keep exploring this.

```{r}
# Filter for Pac-12 teams
pac12_trends <- Stats %>%
  filter(Conference == "Pac-12 Conference") %>%
  mutate(Year = year(ymd(Date)),
         RushingPct = RushingAtt / (PassingAtt + RushingAtt) * 100) %>%
  group_by(TeamFull, Year) %>%
  summarise(avg_rushing_pct = mean(RushingPct, na.rm = TRUE))

# Faceted line chart for each team
ggplot(pac12_trends, aes(x = Year, y = avg_rushing_pct, color = TeamFull)) +
  geom_line(size = 1) +
  facet_wrap(~ TeamFull, scales = "free_y") +  # Create a separate plot for each team
  labs(title = "Rushing Percentage Over Time by Pac-12 Teams",
       x = "Year", y = "Average Rushing Percentage") +
   scale_y_continuous(limits = c(25, 80))+
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend as it’s redundant with facets

   

```





